Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.00

### 2019-02-21 manuscript.Rmd
text updates

## EiaD 2019 demonstrates that organoids, stem cell-derived, and fetal tissue retina have closely related transcriptomes
There are, currently, three approaches to studying developing human retina: 1. post-mortem fteal tissue, 2. organoids, and 3. stem cell-derived cells. As an example of how the EiaD 2019 dataset can be used to address pan-study questions, we looked at whether at how well these three approaches to studying developing retina compare at a transcriptomic level. Should organoid and stem cell-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.

To evaluate this question, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the four types of retina tissue (adult, fetal, iSPC, and organoid) (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`). Here we see three distinct groupings: adult retina (2), developing retina (3), and stem cell/undifferentiated (1). We do see several organoid tissues in cluster 1, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X'). All of the stem cell-based retina samples are older than 50 days.

@@

tsne_50_prep %>% 
  rowwise() %>% 
  mutate(Age = suppressWarnings(as.integer(str_split(sample_attribute, '_')[[1]][2]))) %>% 
  mutate(Origin = as.character(Origin)) %>% 
  mutate(Origin = case_when(Age < 30 ~ 'Organoid < 30 days',
                            Age > 30 ~ 'Organoid > 30 days',
                            TRUE ~ Origin)) %>% 
  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days')),
         Tissue = case_when(sample_accession == 'SRS1572426' ~ 'Retina',
                            TRUE ~ Tissue)) %>% 
  mutate(ZoomTissue = case_when((Tissue == 'Retina' & Sub_Tissue != 'Retina - Adult Tissue') | Tissue == 'ESC' ~ 'Early Retina',
                                TRUE ~ Sub_Tissue),
         Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
                           TRUE ~ Tissue),
         Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label)) %>% 
  mutate(Group = factor(Group, levels = c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE','GTEx'))) %>% 
  ggplot(aes(x=X1,y=X2)) +
  scale_shape_manual(values=c(0:20,35:50)) +
  #geom_point(size=8, alpha=0.2, aes(colour=Group)) +
  geom_point(size=4, alpha=1, aes(shape=Origin, colour = Group)) +
  xlab('t-SNE 1') + ylab('t-SNE 2') +
  facet_zoom(xy = Group == 'Retina', zoom.size = 2, horizontal = F, zoom.data=zoom) + 
  geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 10) + 
  stat_chull(data=tsne_zoom %>% 
               mutate(`Sub-Tissue Cluster` = Cluster_Tissues),aes(fill=`Sub-Tissue Cluster`), alpha = 0.3) +
  scale_color_discrete_sequential(palette = 'viridis') + 
  scale_fill_viridis_d(option = 'plasma') +
  theme_bw() +
  guides(colour = guide_legend(override.aes = list(alpha = 1), ncol=2),
         shape = guide_legend(ncol =2 )) +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        text = element_text(family = 'Linux Libertine O', size = 6))

@@

>>>>

## EiaD 2019 demonstrates that organoids, stem cell-derived, and fetal tissue retina have closely related transcriptomes
There are, currently, two major approaches to studying developing human retina: post-mortem fetal tissue and stem-cell based organoids. We looked at whether at how well these approaches to studying developing retina compare at a transcriptomic level, both for tissue - organoid relationships and how well they correlate across early development.

To evaluate how the tissues and organoids compare at a transcriptome level, we look at the same t-SNE plot from `r fig_cap(name = 'tSNE', display = 'cite')` and focus in on the three types of retina tissue (adult, fetal, and organoid) (`r fig_cap(name = 'tsneFetalRetina', display = 'cite')`). Here we see three distinct groupings: adult retina (2), developing fetal retina and stem cell organoid (3), and undifferentiated and early differentiating stem cells (1). We do see several organoid tissues in cluster 1, but these share one important difference from the rest of the organoid samples in cluster 3: they have been differentiating for less than 30 days (shape 'X'). All of the stem cell-based retina samples in cluster 2 are older than 50 days.

@@

zoom_plot <- tsne_50_prep %>% 
  rowwise() %>% 
  mutate(Age = suppressWarnings(as.integer(str_split(sample_attribute, '_')[[1]][2]))) %>% 
  mutate(Origin = as.character(Origin)) %>% 
  mutate(Origin = case_when(Age < 30 ~ 'Organoid < 30 days',
                            Age > 30 ~ 'Organoid > 30 days',
                            TRUE ~ Origin)) %>% 
  mutate(Origin=factor(Origin, levels=c('Adult Tissue', 'Fetal Tissue', 'Stem Cell', 'Cell Line', 'Organoid < 30 days', 'Organoid > 30 days')),
         Tissue = case_when(sample_accession == 'SRS1572426' ~ 'Retina',
                            TRUE ~ Tissue)) %>% 
  mutate(ZoomTissue = case_when((Tissue == 'Retina' & Sub_Tissue != 'Retina - Adult Tissue') | Tissue == 'ESC' ~ 'Early Retina',
                                TRUE ~ Sub_Tissue),
         Group = case_when(study_accession == 'SRP012682' ~ 'GTEx',
                           TRUE ~ Tissue),
         Label = case_when(grepl('^0', Label) ~ '', TRUE ~ Label)) %>% 
  mutate(Group = factor(Group, levels = c('Cornea', 'ESC', 'Lens', 'Retina', 'RPE','GTEx'))) %>% 
  ggplot(aes(x=X1,y=X2)) +
  scale_shape_manual(values=c(0:20,35:50)) +
  #geom_point(size=8, alpha=0.2, aes(colour=Group)) +
  geom_point(size=4, alpha=1, aes(shape=Origin, colour = Group)) +
  xlab('t-SNE 1') + ylab('t-SNE 2') +
  facet_zoom(xy = Group == 'Retina', zoom.size = 2, horizontal = F, zoom.data=zoom) + 
  geom_label_repel(data=tsne_zoom,size = 2, aes(label=Label), alpha=0.7, box.padding = unit(0.3, "lines"), force = 10) + 
  stat_chull(data=tsne_zoom %>% 
               mutate(`Sub-Tissue Cluster` = Cluster_Tissues),aes(fill=`Sub-Tissue Cluster`), alpha = 0.3) +
  scale_color_discrete_sequential(palette = 'viridis') + 
  scale_fill_viridis_d(option = 'plasma') +
  theme_bw() +
  guides(colour = guide_legend(override.aes = list(alpha = 1), ncol=2),
         shape = guide_legend(ncol =2 )) +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        text = element_text(family = 'Linux Libertine O', size = 6))

@@

Furthermore we also
